\thesistitle

### 2025-07-22 manuscript/part1.tex
correction reading 1: streamline, comprehension and mistakes for foundation

Created by the supervising team - whose research spans formal verification, model checking, temporal logics and automata - the MiniProb DSL provides a concise notation
for probabilistic process models\cite{guideMiniProb}. At present, MiniProb tooling is limited to a Haskell-based parser, yielding a tedious workflow: source code is written
in a generic editor and syntax errors discovered one at a time as the parser aborts on its first failure. A modern editor extension - for example, a
Visual Studio Code plugin providing syntax highlighting and live diagnostics - would dramatically streamline development. Such an interface lowers the barrier to entry for
new users and also offers a familiar environment in which MiniProb programs can be authored and the further automatically converted into automata for formal verification \cite{2025modelcheckingprobabilisticoperator,POPACheck}.
In the context of cyber-physical systems - where interconnected devices underpin critical infrastructure - and the ever-growing complexity and threat landscape of cyberspace,
integrated verification support is essential to ensure system reliability, preempt security vulnerabilities, and uphold correctness guarantees under rigorous temporal and
probabilistic constraints.

\label{sec:scope}
\section{Objectives and Scope}
The main goal of the thesis was to create a \textit{Language Extension} ?NAME? for \textit{MiniProb}. Language extensions are a big part of
VS Code's extension ecosystem and enable the editor to utilize custom language tooling. Such extensions support the user during the implementation process
while writing code, by providing language assistance like syntax validation or code completion. While this implementation targets VS Code, the underlying
logic and functionality are not limited to it - through the use of the \textit{Language Server Protocol (LSP)}, the developed tooling can be reused across
various editors and IDEs that support the protocol. The core functionality of these extensions stems from \textit{parsers} which, against a formal language
definition, convert written text into abstracted parts \ref{sec:parser}.

In this thesis, a generative context-free grammar is used to formally describe \textit{MiniProb} \ref{sec:miniprob}, a domain-specific language (DSL) designed for authoring \textit{POMC} files.
With this grammar as a foundation, parsers enable the implementation of a fully functional language extension to aid developers writing POMC files.\\
Ultimately, the resulting extension, titled ?NAME?, is intended to provide comprehensive language support. This includes syntactic and semantic analysis
through syntax highlighting and validation, accurate resolution of symbol references and rudimentary code completion, as well as implemented type checking mechanisms
to ensure compatibility during development.

@@

Finally, the complete source code is hosted on GitHub at \url{https://github.com/e12224207/miniprob}, and prebuilt \code{.vsix} packages can be downloaded from the Releases page for straightforward installation.
This documentation serves as the definitive guide to the application developed in that repository, underpinned by the theoretical foundations of service customization and language extension principles.

@@

\label{sec:formallang}
\section{Formal Language Theory}

@@

\textbf{Type-3} grammars, or \emph{regular} grammars, generate exactly the class of regular languages. Their productions are restricted to the two equivalent styles:
\[
  right-reg.: \quad A \;\to\; aB \quad or \quad A \;\to\; a
  \qquad and \qquad
  left-reg.: \quad A \;\to\; Ba \quad or \quad A \;\to\; a,
\]
where \(A,B\) are nonterminals and \(a\) is a terminal. Regular expressions provide an alternative, declarative notation for these languages - each expression
can be transformed into a regular grammar, and vice versa. Further, every grammar in the Chomsky hierarchy can be characterized by an equivalent acceptor automaton, which admits all input that ends in valid states as words of the language.
In the case of Type-3 grammars, this correspondence yields deterministic or nondeterministic finite automata. Leveraging these conversions makes regular grammars
invaluable in practice (for example, in lexical analysis, pattern matching, and protocol verification), even though their expressive power is the most limited.
\textbf{Type-2} grammars, or \emph{context-free} grammars (CFGs), produce all context-free languages.
Their rules take the general form  \[A \;\to\; \alpha\] where \(A\) is a single nonterminal and \(\alpha\) is any string of terminals and nonterminals.
This additional flexibility captures nested, hierarchical structures-like balanced parentheses or most programming-language syntaxes - that regular grammars cannot.
Each CFG is accepted by a corresponding pushdown automaton: the grammar expansions map naturally onto push and pop operations on the PDA's stack, making the grammar-automaton
equivalence at this level another cornerstone of formal language theory.

@@

\subsection{Backus-Naur Form (BNF) and Variants}
Even though production rules, alphabet specifications, and automaton definitions suffice to unambiguously define the set of valid strings/sentences in a language, their notation
tends to be opaque and cumbersome in practice, prompting interest in forming intuitively comprehensible grammar definitions. The Backus-Naur Form (BNF),
created by John Backus with contributions by Peter Naur and released in their Algol-60 report\cite{ALGOL60}, was a pivotal introduction furnishing a concise, human-readable syntax for language generation.
By expressing each rule as
\[
  \langle\mathrm{nonterminal}\rangle \;::=\; expansion_1 \;|\; expansion_2 \;|\;\dots
\]
BNF allows both sequencing and alternation of multiple expansions, enabling language designers to articulate complex, nested structures while preserving structural
clarity and high readability.
This declarative approach not only supports rigorous specification but also facilitates mechanical parsing, thereby advancing the practice of compiler development and
language tooling.

While BNF provides a clear, formal way to describe language syntax, it suffers from verbosity and redundancy - pure BNF grammars often become bloated when encoding
common patterns like optionals or repetitions requiring auxiliary nonterminals for a sufficient description and, over time, has been extended and modified for various
use-cases spawning a new family of grammar notations. The \textbf{Extended Backus-Naur Form} extends BNF by adding more expressive meta-syntax, introducing operators
similar to \textit{Regular Expressions} allowing the use of optional or repeatable expressions, commonly indicated by brackets but not limited to '[\dots]' and '\{\dots\}',
and the use of comments. Such extensions make grammars more compact and easier to read without changing the fundamental class of languages they describe.
Multiple distinct instances of EBNF's have been development, all with minor syntactic differences, yet there really is no on-for-all EBNF used in practice, although
a standardized ISO/IEC 14977 version exists. \cite{jinks2004bnf,jinks2004ebnfvariants}\\
Within the scope of this thesis, the Extended Backus-Naur Form (EBNF) is particularly significant, since Langium employs its own custom EBNF dialect \ref{sec:langium-grammar}
to define the grammars underpinning its language-assistance features. \ref{sec:langium-grammar}

@@

The core functionality of language extensions stems from \textit{parsers} which, against a formal language
definition, convert written text into abstracted parts, validating the code in the process. These parts can then be used to extend the capabilities of the
tooling to encompass referential validation, type checking, code completion, diagnostics, and other language services that enhance the development experience.
In order for parsers to function accordingly, they require formal language definitions or models. These definitions range from various types of grammars to
automata such as PDAs, which provide the structural and syntactic rules necessary for correct interpretation and validation of source code.
Based on these definitions, parsers are able to systematically identify the hierarchical structure of a program by recognizing patterns, token sequences, and nested
constructs, ensuring that the code adheres to the expected form before any processing or analysis of the code occurs.
With a defined grammar as a foundation, parsers enable the implementation of fully functional development tools - for example, a language extension that assists with code editing.
\\

@@

Top-down parsers, such as those following the \textit{LL($k$)} model, process input from left to right, constructing parse trees from the root downward.
These parsers require grammars to avoid \textit{left recursion}, a situation where a non-terminal refers to itself as the first symbol on the right-hand side of a
production (e.g., \texttt{A}~$\rightarrow$~\texttt{A}~$\alpha$). Left-recursive rules can lead to infinite recursion during top-down parsing.
In contrast, \textit{bottom-up} parsers, including \textit{LR($k$)} and its variants, build parse trees from leaves to root and can naturally accommodate
left-recursive structures.

@@

When designing programming languages, the abstract syntax is of particular interest, as its streamlined view of the language is ideally suited for developing
validation constraints or type systems and for general interpretation. This is done most commonly by encoding the syntax into a \textbf{Abstract Syntax Tree (AST)},
a tree-like structure containing nodes for each high-level construct (such as expressions, declarations, or statements) connected according to their logical hierarchy,
while deliberately omitting lexical details to yield a canonical. \cite{slonneger1995specifying}
The \textbf{Concrete Syntax Tree (CST)} is the specialized counterpart to the AST and consists of a full parse-tree, preserving every terminal and nonterminal
token(keywords, delimiters, etc.) and mirrors each grammar production in its node structure, providing the complete syntactic context possibly needed for precise error
reporting and tooling. \cite{aho2006compilers}

@@

Imperative languages, in contrast to declarative ones, describe \textit{how} a computation will accomplish the desired end-state by following explicit sequences
of commands. These commands consist of statements that actively manipulate the computations state, which is typically managed through assignments to variables
and taking charge of the control flow with provided language constructs (loops, conditionals, calls, procedures, etc.).
Consequently, correct execution is no longer responsibility of the underlying system, but the implementation and therefore the programmer themselves. The imperative paradigm
reflects von Neuman machines - accessing memory and hardware directly. This fined grained control enables the use of low-level optimization techniques \cite{lowLvelOpt}.
As programmers instruct the machine how to do something, certain performance characteristics and resource constraints can be impelled depending on the context of the computation.
Such techniques encompass memory and cache management, \textit{Instruction-Set} optimization or compiler optimization.

However, while this explicit control provides developers with great functional freedom, it also places a greater burden on them to manage every detail of execution.
Because the programmer must specify each step - from how data is stored in memory to the order in which operations occur - imperative code can become verbose and cluttered obscuring the high-level intent.
Moreover, mutable state and side effects introduce the risk of subtle bugs, especially in concurrent or parallel settings where unintended interactions between state changes
can lead to race conditions. Also, formal reasoning and verification are more complex, since proving correctness requires tracking every state
transition rather than relying on \textit{referentially transparent} expressions.

Typical use cases for imperative programming include algorithmic or process-oriented problems where an explicit series of steps is natural.
Low-level system programming, embedded development, and performance-sensitive routines frequently rely on imperative constructs to manage hardware resources directly.More actions
In scenarios that demand fine-grained stateful interactions - such as updating user interfaces incrementally, reading and writing files in a specific order, or
controlling physical systems via commands - imperative languages shine by giving developers precise command over execution.
Simulation engines, game logic, and scripting or “recipe”-style automation languages also favor imperatively describing each step to achieve the desired outcome.

@@

The close relationship of DSL and domain furnishes multiple advantages over conventional GLPs: it reduces the amount of general programming knowledge needed,
expanding the pool of developers from traditional programmers to also domain experts. Moreover, DSLs declutter code, rather than describing a subproblem of the domain
with multiple lines of code, fewer are needed because constructs inherently carry domain specific information, also contributing to broader comprehensibility for those familiar with the domain.
DSLs solutions are also easier to formally verify, as they are restricted to a limited context, than GLPs for which proving correctness is extremely hard.

However, DSLs can suffer from limited scope and flexibility even within their target domain, forcing developers to fall back on general-purpose languages when they need functionality outside the
DSL's narrow focus. Furthermore, if the domain itself is complex or rapidly evolving, creating and maintaining a DSL may still demand deep domain expertise, because the abstractions must
accurately capture all necessary domain rules and nuances.

@@

Whichever paradigm a DSL follows is mostly guided by the specifics of the domain to describe. Even though DSLs usually adhere to a declarative approach \cite{sigplanDSL}, which follows from a
major motivation: wanting to describe \textit{what} is expected of the domain in high-level abstraction, a domain may emit qualities that invoke more imperative thinking.
If the nature of a domain is inherently composed of sequential actions, consists of procedures or stateful operations, the imperative paradigm is more appropriate.
Imperative DSLs require the user to script solutions in step-by-step fashion using domain-specific commands and statements. Essentially,the user embeds domain concepts as a sequential programming model,
where the control flow and state transitions of the domain computations have to be managed.

In domains where fine-grained control and explicit ordering matter, imperative DSLs provide maximal flexibility, because every aspect of the execution is adjustable.
This level of control becomes crucial when timing or order of operations yield different outcomes, or when fine-tuned performance optimizations are required.
Furthermore, building an applicable imperative DSL is often simpler than creating a declarative one: an interpreter processes each command in sequence,
or a translator converts DSL instructions into a host general-purpose language.

@@

Cyber-Physical System (CPS) is an umbrella term broadly referring to any systems which tightly integrate both computational logic (cyber) and physical components, allowing interplay
through designated interfaces. \cite{cpsModelDesc} The two parts of a CPS have to be seamlessly interconnected enabling efficient communication and consistent information
relying?? \cite{cpsChallengesAndFuture}, as both sides heavily depend on one another.
The cyber-physical relationship can be generally describes by three roles \textit{controller/agent}, \textit{sensors} and \textit{activators},
which are connected by an underlying network.
Controllers, representing the 'cyber', are occupied with computation and relying fitting instructions based on te observational data provided by the physical sensors.
Sensor and Actuators embody the 'physical' and are the ones interacting with the environments in which the system resides. A sensors jobs is to gather intel from its
surroundings communicating it in order for the agent to make correct decisions and predictions and to monitor the system. Without sufficient data, the controller is left
blind and possibly  unable to perform as intended. While sensors passively interact with the environment by observing, actuators do so actively by altering the systems state
- carrying out tasks instructed by the controller. This sense->compute->actuate cycle is executed iteratively, forming a continuous feedback loop that enables the system to
respond dynamically to changes in its physical environment.

@@

CP-Systems are inherently multidisciplinary structures intersecting engineering domains (mechanical, electrical, civil, etc.) with computer science.
Because these domains employ differing models for ... , effectively combining them is non-trivial. Challenges arise during development and operation of CPSs for which
finding solutions is imperative. \cite{cpsChallengesAndFuture} First and foremost is the concern of \textbf{reliability and safety}. Occurrences of failures can have dire real-world
consequences with catastrophic end results. This means systems must handle unexpected conditions gracefully and robustly. But ensuring safety through extensive verification
and validation is complicated by the combination of continuous physical dynamics and discrete decision logics. Another challenge is \textbf{real-time performance}, as computing
elements must keep pace with the physical processes, which usually require low-latency responsiveness including the network. Since CPSs are used in infrastructure and other
critical sites that may be targets of cyber attacks, guarantying a high standard of \textbf{Security} is essential. Additionally, sensors are not completely reliable -
possibly introducing noise to measurements - and environmental conditions can vary drastically leading to \textbf{uncertainty}, which a CPS has to cope with to be resilient. \cite{cpsProbabilisticRobotics}

@@

CPSs are often deployed in safety-critical environments where utmost importance is placed on reliable operation and functional correctness. Yet reaching high levels of confidence
in a systems behavior remains extremely inside cyber-physical domains.\cite{formalMethodsCPSCritical} This difficulty arises from the complex interactions of all the individual integral
components of the system, which can not be effectively tested only employing traditional methods. The inherent entanglement of these systems warrants the use of
specialized procedures that are able to cover and validate the acting components. Unlike conventional simulation based testing, \textit{formal methods} are able to prove correctness against
formal specifications, thereby providing more reliable results and eliminating potential error sources, such as race conditions or violations of safety properties(unnoticed states).

@@

Model checking is of most interest in this case, as the MiniProb DSL \ref{sec:miniprob} was developed with this context in mind. \cite{POPACheck}
It statically verifies a system by exhaustively exploring the state-space and is considered one of the most successful verification techniques. \cite{formalMethodsCPSCritical}
The thorough analysis of reachable states, combined with specifications which are expressible in various forms of modal logic,
offer a strong expressive foundation for property verification \cite{modelCheckingPrinceples}. However, the systematic exploration of states leads to poor performance, especially for model with
complex and big state spaces. Moreover, as most model checking implementation employ automata-based solutions, systems are usually converted into
equivalent state machines which can result in an explosion of states.\cite{}
Model checking is generally only used in conjunction with finite-state machines for aforementioned reasons. Still, theres been advances trying to apply it to infinite state spaces.
\cite{modelCheckingInfCounter,modelCheckingInfSymb}. The resulting verdicts for infinite-state models are typically approximations or are restricted by bounded analysis,
such as checking only a finite number of steps or abstracting unbounded data domains. As a result, the verification outcomes may not provide absolute guarantees for all
possible executions, but rather assurances that hold under specific assumptions, abstractions, or within predefined limits of a system's behavior.

@@

Probabilistic programming has multiple fitting applications within CPSs, especially where reasoning under uncertainty is essential. Generally, probabilistic programming
allows controllers to plan even in unpredictable environments and settings.

\begin{multicols}{2}
  \begin{itemize}
    \item When measurements contain noise or are incorrect, PPL models can still be applied to extract useful information. Instead of treating the raw reading of a sensor as the “truth,”
  the readings are considered samples for a \textit{sensor’s model}. Not only is it possible to reduce the noise in the measurements by calculating the posterior,
  but those noise characteristics are also used to flesh out the predictive model itself.
  This enables PPLs to adaptively adjust noise models online when, for example, sensors deteriorate over time.
  Additionally - with most CPSs employing multiple sensors - a PPL naturally supports \textit{sensor fusion}, \cite{cpsSensorFusion} combining data from heterogeneous sources
  to produce a coherent estimate. By defining each sensor's observation as a conditional distribution, a probabilistic program
  weights each reading according to its uncertainty: more reliable sensors contribute more heavily, while inconsistent or faulty readings are down-weighted or flagged.
  Inference then yields a joint posterior over the true state, allowing the system to reconcile disagreements between sensors, detect potential sensor failures by inferring
  latent bias variables, and maintain a quantified confidence.
  \item Which leads us into robotics, a quintessential cornerstone of CPS. Instead of working on raw observational data and making assumptions in incomplete cases, robots can adopt
  a \textit{belief} based on an observational model \cite{cpsProbabilisticRobotics}. The observational model couples motion dynamics (how actions translate into state
  transitions) with existing sensor models, providing a full posterior distribution that the robot can use to imply different levels of confidence for its actions.
  Furthermore, robots often operate in changing environments or with components that degrade over time. Within a PPL, unknown parameters, such as friction
  coefficients or sensor calibration offsets, can be treated as latent variables and perform Bayesian learning to infer them from incoming data. The same inference mechanism used for state
  estimation thus also adapts the model online, continuously refining both motion dynamics and sensor mappings.

>>>>

The \textbf{MiniProb} DSL, created by the supervising team, provides a concise notation for probabilistic process models\cite{guideMiniProb}. At present,
MiniProb tooling is limited to a Haskell-based parser, yielding a tedious workflow: source code is written
in a generic editor and syntax errors discovered one at a time as the parser aborts on its first failure. A modern editor extension - for example, a
Visual Studio Code plugin providing syntax highlighting and live diagnostics - would dramatically streamline development. Such an interface lowers the barrier to entry for
new users and also offers a familiar environment in which MiniProb programs can be authored and then further automatically converted into automata for formal verification \cite{2025modelcheckingprobabilisticoperator,POPACheck}.
In the context of cyber-physical systems - where interconnected devices underpin critical infrastructure - and the ever-growing complexity and threat landscape of cyberspace,
integrated verification is essential to ensure system reliability, preempt security vulnerabilities, and uphold correctness guarantees under rigorous specifications.

\label{sec:scope}
\section{Objectives and Scope}
The main goal of the thesis was to create a \textit{Language Extension} for \textit{MiniProb}. Language extensions are a big part of
VS Code's extension ecosystem and enable the editor to utilize custom language tooling. Such extensions support the user during the implementation process
while writing code, by providing language assistance like syntax validation or code completion. While this implementation targets VS Code, the underlying
logic and functionality are not limited to it - through the use of the \textit{Language Server Protocol (LSP)}, the developed tooling can be reused across
various editors and IDEs that support the protocol. The core functionality of these extensions stems from \textit{parsers} which, against a formal language
definition (\ref{sec:formallang}), convert written text into abstracted parts (\ref{sec:parser}).

In this thesis, a generative context-free grammar is used to formally describe \textit{MiniProb} \ref{sec:miniprob} and a domain-specific language (DSL) is
designed (\ref{sec:langium-grammar}), for authoring \textit{POMC} files. With a formal language definition as foundation, parsers enable the implementation of a fully functional
language extension to aid developers writing POMC files.\\
Ultimately, the resulting extension, titled ?NAME?, is intended to provide comprehensive language support. This includes syntactic and semantic analysis
through syntax highlighting and validation, accurate resolution of symbol references and rudimentary code completion, as well as implemented type checking mechanisms
to ensure compatibility during development.

@@

Finally, the complete source code is hosted on GitHub at \url{https://github.com/e12224207/miniprob}, and prebuilt \code{.vsix} packages can be downloaded from the Releases page for straightforward installation.
This documentation serves as the definitive guide to the application developed in that repository, underpinned by theoretical foundations to contextualize the themes of this thesis.

@@

\section{Formal Language Theory}
\label{sec:formallang}

@@

\textbf{Type-3} grammars, or \emph{regular} grammars, generate exactly the class of regular languages. Their productions are restricted to the two equivalent styles:
\[
  right-reg.: \quad A \;\to\; aB \quad or \quad A \;\to\; a
  \qquad and \qquad
  left-reg.: \quad A \;\to\; Ba \quad or \quad A \;\to\; a,
\]
where \(A,B\) are nonterminals and \(a\) is a terminal. Regular expressions provide an alternative, declarative notation for these languages - each expression
can be transformed into a regular grammar, and vice versa. Further, every grammar in the Chomsky hierarchy can be characterized by an equivalent acceptor automaton, which admits all input that ends in valid states as words of the language.
In the case of Type-3 grammars, this correspondence yields deterministic or nondeterministic finite automata. Leveraging these conversions makes regular grammars
invaluable in practice (for example, in lexical analysis, pattern matching, and protocol verification), even though their expressive power is the most limited.
\textbf{Type-2} grammars, or \emph{context-free} grammars (CFGs), produce all context-free languages.
Their rules take the general form  \[A \;\to\; \alpha\] where \(A\) is a single nonterminal and \(\alpha\) is any string of terminals and nonterminals.
This additional flexibility captures nested, hierarchical structures - like balanced parentheses or most programming-language syntaxes - that regular grammars cannot.
Each CFG is accepted by a corresponding pushdown automaton: the grammar expansions map naturally onto push and pop operations on the PDA's stack, making the grammar-automaton
equivalence at this level another cornerstone of formal language theory.

@@

\subsection{Backus-Naur Form (BNF) and Variants}
Even though production rules, alphabet specifications, and automaton definitions suffice to unambiguously define the set of valid strings/sentences in a language, their notation
tends to be opaque and cumbersome in practice, prompting interest in forming intuitively comprehensible grammar definitions. The Backus-Naur Form (BNF),
created by John Backus with contributions by Peter Naur and released in their Algol-60 report\cite{ALGOL60}, was a pivotal introduction furnishing a concise, human-readable syntax for language generation.
By expressing each rule as
\[
  \langle\mathrm{nonterminal}\rangle \;::=\; expansion_1 \;|\; expansion_2 \;|\;\dots
\]
BNF allows both sequencing and alternation of multiple expansions, enabling language designers to articulate complex, nested structures while preserving structural
clarity and high readability.
This declarative approach not only supports rigorous specification but also facilitates parsing, advancing the practice of compiler development and
language tooling.

While BNF provides a clear, formal way to describe language syntax, it suffers from verbosity and redundancy - pure BNF grammars often become bloated when encoding
common patterns like optionals or repetitions requiring auxiliary nonterminals for a sufficient description, and, over time, has been extended and modified for various
use-cases spawning a new family of grammar notations. The \textbf{Extended Backus-Naur Form} extends BNF by adding more expressive meta-syntax, introducing operators
similar to \textit{Regular Expressions} allowing the use of optional or repeatable expressions, commonly indicated by brackets but not limited to '[\dots]' and '\{\dots\}',
and the use of comments. Such extensions make grammars more compact and easier to read without changing the fundamental class of languages they describe.
Multiple distinct instances of EBNF's have been development, all with minor syntactic differences, yet there really is no on-for-all EBNF used in practice, although
a standardized ISO/IEC 14977 version exists. \cite{jinks2004bnf,jinks2004ebnfvariants}\\
Within the scope of this thesis, the Extended Backus-Naur Form (EBNF) is particularly significant, since Langium employs its own custom EBNF dialect (\ref{sec:langium-grammar})
to define the grammars underpinning its language-assistance features.

@@

The core functionality of language extensions stems from \textit{parsers} which, against a formal language
definition, convert written text - most often program code - into abstracted parts. These parts can then be used to extend the capabilities of the
tooling to encompass referential validation, type checking, code completion, diagnostics, and other language services that enhance the development experience.
In order for parsers to function accordingly, they require formal language definitions or models. Such specifications range from various types of grammars to
automata, and provide the structural and syntactic rules necessary for correct interpretation and validation of the source code.
Based on these definitions, parsers are able to systematically identify the hierarchical structure of a program by recognizing patterns, token sequences, and nested
constructs, ensuring that the code adheres to the expected form before any further processing or analysis of it occurs.
\\

@@

Top-down parsers, such as those following the \textit{LL($k$)} model, process input from left to right, constructing parse trees from the root downward.
These parsers require grammars to avoid \textit{left recursion}, a situation where a non-terminal refers to itself as the first symbol on the right-hand side of a
production (e.g., \texttt{A}~$\rightarrow$~\texttt{A}~$\alpha$). Left-recursive rules can lead to infinite recursion during top-down parsing.
In contrast, \textit{bottom-up} parsers, including \textit{LR($k$)} and its variants still read left to right but build parse trees from leaves to root and can naturally accommodate
left-recursive structures.

@@

When designing programming languages, the abstract syntax is of particular interest, as its streamlined view of the language is ideally suited for developing
validation constraints or type systems and for general interpretation. This is done most commonly by encoding the syntax into a \textbf{Abstract Syntax Tree (AST)},
a tree-like structure containing nodes for each high-level construct (such as expressions, declarations, or statements) connected according to their logical hierarchy,
while deliberately omitting lexical details. \cite{slonneger1995specifying}
The \textbf{Concrete Syntax Tree (CST)} is the specialized counterpart to the AST and consists of a full parse-tree, preserving every terminal and nonterminal
token(keywords, delimiters, etc.) and mirrors each grammar production in its node structure, providing the complete syntactic context possibly needed for precise error
reporting and tooling. \cite{aho2006compilers}

@@

Imperative languages, in contrast to declarative ones, describe \textit{how} a computation will accomplish the desired end-state by following explicit sequences
of commands. These commands consist of statements that actively manipulate the computations state, which is typically managed through assignments to variables
and taking charge of the control flow with provided language constructs (loops, conditionals, calls, procedures, etc.).
Consequently, correct execution is no longer responsibility of the underlying system, but the implementation and therefore the programmer themselves. The imperative paradigm
reflects von Neuman machines: accessing memory and hardware directly. This fined grained control enables the use of low-level optimization techniques \cite{lowLvelOpt}.
As programmers instruct the machine how to do something, certain performance characteristics and resource constraints can be shifted in and out of focus depending on the context of the computation.
Such techniques encompass memory and cache management, \textit{Instruction-Set} optimization or compiler optimization.

However, while this explicit control provides developers with great freedom, it also places a greater burden on them to manage every detail of the execution.
Because the programmer must specify each step - from how data is stored in memory to the order in which operations occur - imperative code can become verbose and cluttered obscuring the high-level intent.
Moreover, mutable state and side effects introduce the risk of subtle bugs, especially in concurrent or parallel settings where unintended interactions between state changes
can lead to race conditions. Also, formal reasoning and verification are more complex, since proving correctness requires tracking every state
transition rather than relying on \textit{referentially transparent} expressions.

Typical use cases for imperative programming include algorithmic or process-oriented problems where an explicit series of steps is natural.
Low-level system programming, embedded development, and performance-sensitive routines frequently rely on imperative constructs to manage hardware resources directly.
In scenarios that demand fine-grained stateful interactions - such as updating user interfaces incrementally, reading and writing files in a specific order, or
controlling physical systems via commands - imperative languages shine by giving developers precise command over execution.
Simulation engines, game logic, and scripting or “recipe”-style automation languages also favor imperatively describing each step to achieve the desired outcome.

@@

The close relationship of DSL and domain furnishes multiple advantages over conventional GLPs: it reduces the amount of general programming knowledge needed,
expanding the pool of developers from traditional programmers to also include domain experts. Moreover, DSLs declutter code, rather than describing a subproblem of the domain
with multiple lines of code, fewer are needed because constructs inherently carry domain specific information, also contributing to broader comprehensibility for those familiar with the domain.
DSL solutions are also easier to formally verify - as they are restricted to a limited context - than GLPs for which proving correctness is extremely hard.

However, DSLs can suffer from limited scope and flexibility even within their target domain, forcing developers to fall back on general-purpose languages when they need functionality outside a
DSL's narrow focus. Furthermore, if the domain itself is complex or rapidly evolving, creating and maintaining a DSL may still demand deep domain expertise, because the abstractions must
accurately capture all necessary domain rules and nuances.

@@

Whichever paradigm a DSL follows is mostly guided by the specifics of the domain to describe. Even though DSLs usually adhere to a declarative approach \cite{sigplanDSL}, which follows from a
major motivation: wanting to describe \textit{what} is expected of the domain in high-level abstraction, a domain may emit qualities that invoke more imperative thinking.
If the nature of a domain is inherently composed of sequential actions, consists of procedures or stateful operations, the imperative paradigm is more appropriate.
Imperative DSLs require the user to script solutions in step-by-step fashion using domain-specific commands and statements. Essentially, the user embeds domain concepts as a sequential programming model,
where the control flow and state transitions of the domain computations have to be managed.

In domains where fine-grained control and explicit ordering matter, imperative DSLs provide maximal flexibility, because every aspect of the execution is adjustable.
This level of control becomes crucial when timing or order of operations yield different outcomes, or when fine-tuned performance optimizations are required.
Furthermore, building an applicable imperative DSL is often simpler than creating a declarative one: an interpreter processes each command in sequence,
or a translator converts DSL instructions into a general-purpose language.

@@

Cyber-Physical System (CPS) is an umbrella term broadly referring to any systems which tightly integrate both computational logic (cyber) and physical components, allowing interplay
through designated interfaces. \cite{cpsModelDesc} The two parts of a CPS have to be seamlessly interconnected enabling efficient communication and consistent information
exchange \cite{cpsChallengesAndFuture}, as both sides heavily depend on one another.
The cyber-physical relationship can be generally describes by three roles \textit{controller/agent}, \textit{sensors} and \textit{activators},
which are connected by an underlying network.
Controllers, representing the 'cyber', are occupied with computation and relying fitting instructions based on the observational data provided by the physical sensors.
Sensor and Actuators embody the 'physical' and are the ones interacting with the environments in which the system resides. A sensors jobs is to gather intel from its
surroundings communicating it in order for the agent to make correct decisions and predictions and to monitor the system. Without sufficient data, the controller is left
blind and possibly  unable to perform as intended. While sensors passively interact with the environment by observing, actuators do so actively by altering the systems state
- carrying out tasks instructed by the controller. This sense->compute->actuate cycle is executed iteratively, forming a continuous feedback loop that enables the system to
respond dynamically to changes in its physical environment.

@@

CPSs are inherently multidisciplinary structures intersecting engineering domains (mechanical, electrical, civil, etc.) with computer science.
Because these domains employ differing models for various tasks, thereby effectively combining them is non-trivial. Challenges arise during development and operation of CPSs for which
finding solutions is imperative. \cite{cpsChallengesAndFuture} First and foremost is the concern of \textbf{reliability and safety}. Occurrences of failures can have dire real-world
consequences with catastrophic end results. This means systems must handle unexpected conditions gracefully and robustly. But ensuring safety through extensive verification
and validation is complicated by the combination of continuous physical dynamics and discrete decision logics. Another challenge is \textbf{real-time performance}, as computing
elements must keep pace with the physical processes, which usually require low-latency responsiveness including the network. Since CPSs are used in infrastructure and other
critical sites that may be targets of cyber attacks, guarantying a high standard of \textbf{Security} is essential. Additionally, sensors are not completely reliable -
possibly introducing noise to measurements - and environmental conditions can vary drastically leading to \textbf{uncertainty}, which a CPS has to cope with to be resilient. \cite{cpsProbabilisticRobotics}

@@

CPSs are often deployed in safety-critical environments where utmost importance is placed on reliable operation and functional correctness. Yet reaching high levels of confidence
in a systems behavior remains extremely challenging inside cyber-physical domains.\cite{formalMethodsCPSCritical} This difficulty arises from the complex interactions of all the individual integral
components of the system, which can not be effectively tested only employing traditional methods. The inherent entanglement of these systems warrants the use of
specialized procedures that are able to cover and validate the acting components. Unlike conventional simulation based testing, \textit{formal methods} are able to prove correctness against
formal specifications, thereby providing more reliable results and eliminating potential error sources, such as race conditions or violations of safety properties(unnoticed states).

@@

\textit{Model checking} is of most interest in this case, as the MiniProb DSL \ref{sec:miniprob} was developed with this context in mind. \cite{POPACheck}
It statically verifies a system by exhaustively exploring the state-space and is considered one of the most successful verification techniques. \cite{formalMethodsCPSCritical}
The thorough analysis of reachable states, combined with specifications which are expressible in various forms of modal logic,
offer a strong expressive foundation for property verification \cite{modelCheckingPrinceples}. However, the systematic exploration of states leads to poor performance, especially for models with
complex and big state spaces. Moreover, as most model checking implementation employ automata-based solutions, systems are usually converted into
equivalent state machines which can result in an explosion of states.\cite{}
Model checking is generally only used in conjunction with finite-state machines for aforementioned reasons. Still, theres been advances trying to apply it to infinite state spaces.
\cite{modelCheckingInfCounter,modelCheckingInfSymb}. The resulting verdicts for infinite-state models are typically approximations or are restricted by bounded analysis,
such as checking only a finite number of steps or abstracting unbounded data domains. As a result, the verification outcomes may not provide absolute guarantees for all
possible executions, but rather assurances that hold under specific assumptions, abstractions, or within predefined limits of a system's behavior.

@@

Probabilistic programming (\ref{sec:pp}) has multiple fitting applications within CPSs, especially where reasoning under uncertainty is essential. Generally, probabilistic programming
allows controllers to plan even in unpredictable environments and settings.

\begin{multicols}{2}
  \begin{itemize}
    \item When measurements contain noise or are incorrect, PPL models can still be applied to extract useful information. Instead of treating the raw reading of a sensor as the “truth,”
  the readings are considered samples for a \textit{sensor's probabilistic model}. Not only is it possible to reduce the noise in the measurements by calculating the posterior,
  but those noise characteristics are also used to flesh out the predictive model itself.
  This enables PPLs to adaptively adjust noise models online when, for example, sensors deteriorate over time.
  Additionally - with most CPSs employing multiple sensors - a PPL naturally supports \textit{sensor fusion}, %\cite{cpsSensorFusion}
  combining data from heterogeneous sources to produce a coherent estimate. By defining each sensor's observation as a conditional distribution, a probabilistic program
  weights each reading according to its uncertainty: more reliable sensors contribute more heavily, while inconsistent or faulty readings are down-weighted or flagged.
  Inference then yields a joint posterior over the "true" state, allowing the system to reconcile disagreements between sensors, detect potential sensor failures by inferring
  latent bias variables, and maintain a quantified confidence.

  \item A PPL expresses system dynamics probabilistically so that sampling yields a distribution of possible futures \cite{cpsPredictiveControl}. By enforcing “chance constraints”
  (e.g., “limit the probability of exceeding a safety threshold to 1\%”), inference or Monte Carlo sampling identifies control inputs that optimize expected performance while
  respecting risk. When the current state is uncertain, planning occurs in belief space - each action triggers a Bayesian update of the state distribution, and planners
  choose sequences that minimize expected cost given that uncertainty. Simultaneously, online Bayesian learning refines model parameters (like friction or drag) over time,
  tightening future predictions; as uncertainty shrinks, control can push closer to limits, whereas rising uncertainty leads to more conservative actions.

  \item Which leads us into robotics, a quintessential cornerstone of CPS. Instead of working on raw observational data and making assumptions in incomplete cases, robots can adopt
  a \textit{belief} based on an observational model \cite{cpsProbabilisticRobotics}. The observational model couples motion dynamics (how actions translate into state
  transitions) with existing sensor models, providing a full posterior distribution that the robot can use to imply different levels of confidence for its actions.
  Furthermore, robots often operate in changing environments or with components that degrade over time. Within a PPL, unknown parameters, such as friction
  coefficients or sensor calibration offsets, can be treated as latent variables and used to perform Bayesian learning to infer them from incoming data. The same inference mechanism used for state
  estimation thus also adapts the model online, continuously refining both motion dynamics and sensor mappings.
